Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-06-01 icass.Rnw
bibliography and cosmetic changes

%\bibliographystyle{apalike}
\bibliographystyle{plainnat}

@@

\title{Long--term perspectives of social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Health and Mortality in Arctic Populations'. The Ninth International
    Congress on Arctic Social Sciences,
    Umeå,
    June 8--12, 2017}}

@@

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken in
this paper. The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013. The outcome is all
cause mortality and cause specific mortality for the categories
cardiovascular diseases and cancers. The analysis of inequality is based on
three social classes, reorganized from Hisclass.

@@

The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Västerbotten in the north of Sweden along the coast of Gulf of Bothnia. This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century. The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s. One living long in memory was the famine of 1867–-1868 \citep{segb14}. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures. The inland was particularly remote and isolated. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden. Diseases reached to places that earlier were spared.

@@

In the data-set analysed here, all individuals 40 years and older ever being resident in either
of the regions are included. The data file contains the variables on social class (HISCLASS) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.

@@

In this paper we analyse mortality differences according to social class based on the classification scheme HISCLASS \citep{lema11}. The only type of socio-economic information available both in Linnaeus and Poplink is occupations. Thus, analysing according to social class is apart from being a scientific choice, also a practical one. Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid. Thus, the social class in the census 1990 is considered to be kept until 2001. This strategy is also used for other cases of missing social class. Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65. For the Linnaeus data, we use a script recoding HISCO into HISCLASS, provided to us by Jonas Helgertz, Lund University.

@@

HISCLASS, the classification system used here, is a ”. . . HISCO-based
historical international social class scheme” \citep{lema11, lema02}. The different classes in HISCLASS represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector. Implicitly it reflects large differences in what is in the forefront of other
categorisations, as for example access to economic and other resources, status, social power et cetera. We have merged the original 12 classes in HISCLASS, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:

\begin{enumerate}
\item {\tt elite}, HISCLASS 1 and 2. Higher managers, professionals and higher white collar.
\item {\tt middle}, HISCLASS 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
\item {\tt worker}, HISCLASS 7, 9, 10, 11 and 12. Workers including farm workers.
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

@@

\subsection{Cause of death}
In the present study, we analyse mortality in the two most common disease
groups in the ages we are investigating -- cardiovascular diseases and
cancers. National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year. Classification systems have however changed. Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system. In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.

@@

The important explanatory variable, {\sc Hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\sc Hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are presented \emph{graphically}. Tables with results from the different models are presented at the end of the article, but are not commented on in detail. The comments on the results are primarily based on the figures and focuses on the central variable in our study -- {\sc Hisclass}. As can be seen from the tables, most variables in the model are significant at 5 per cent level.

@@

\section{The effect of Hisclass, 1851--1950, 1961--2013}

@@

From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
The most interesting feature is the three-way interaction with sex, period, and hisclass.

@@

As before, models are fitted separately for sex and each ten-year time period, from  1901 to
2013. We stratify on HISCLASS and use the covariates urban/rural and {\tt   civst}. The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.

@@

While we assume clear social differences in cardiovascular diseases, this is not necessarily an obvious assumption when it comes to cancers. Different forms of cancers have very different origin and different pathways to disease and death. The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers. In our analyses, there is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated by the already mentioned large proportion of unknown or vague causes at that time. We believe however that cancers were better diagnosed than cardiovascular deaths. This is supported by the stability in proportion of cancer deaths between the two first periods, 12,5 per cent 1901--1925 and 13,5 per cent for the period 1926--1950.

@@

Figure~\ref{fig:12cumsw} shows relative mortality levels according to HISCLASS during the different studied periods for the age group 40--64. Cancer mortality among the élite women was lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods. Consequently, we find a more consistent advantage of higher class in cancers compared to cardiovascular diseases for women. We do however not find the same pattern for men. It is difficult to identify any general features of 20th century male cancer mortality when it comes to social differences.

@@

@@

@@

<<tableincprheart,results='asis'>>=
ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.", 
    label = "tab:tableincpr", digits = 4)
@

@@

@@

The authors of this paper have scrutinised different aspects of the social
pattern of mortality in historical northern Sweden for some time, focusing on
different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}. When it comes to infant and child mortality pre-- and early--transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}. Studies based on microdata on the adult population in history have been rare, mainly due to lack of adequate data. The results from such studies show other patterns than that of infant and child mortality. The expected social gradient in
survival, meaning that people from higher social classes always had an
advantage, was in many cases, particularly among adult men, not found
despite the well-documented socio-economic differences \citep{edli11, edbr12}. There were however large differences between men and women and for different age groups. These earlier studies have suggested that distinct and consistent differences to the advantage of higher classes developed somewhere during and probably quite early in the 20th century \citep{bevp11}. Until now we have however not been able to investigate the continuous development on mortality differences among the adult and elderly
population due to no available data. This is the first time it has been possible, and the results continue to surprise us. The consistent advantage for higher social classes do not appear until quite late in the regions we have analysed. The social pattern of female mortality are in line to what is
expected, but a social gradient in male adult mortality seems to be a recent phenomenon. Furthermore, this transition and the class pattern of mortality differ depending on age group. Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly.

@@

In this paper we have analysed social inequality in health from the
perspective of social classes using HISCLASS. However, there are
alternative ways to analyse social inequality, and we cannot exclude that
such analyses gives other results or at least would modify the image
presented here. In many modern studies categorisations based on income, wealth
or education are common for studies of inequality. They represent different dimensions of 
inequality. One problem is however that reliable data on income and
education are rarely available in historical population databases. 
As a test, we have included analyses of the associations of
education and income to mortality for recent decades. For women as
well as for men, for both age groups and for all-cause mortality as well as
cause-specific mortality we find consistent differences with higher
educational level and more income leading to lower mortality, thus basically confirming what we find in the class analysis. We do not know if education and income are equally important in the past.

@@

This work was undertaken within the Centre for Demographic and Ageing
Research (CEDAR) at Umeå University and with support from the Swedish
Research Council (VR) (grant no. 421-2012-892) and the project Old age, health and inequality: mortality from a life course perspective in Sweden 1986-2009, financed by the Swedish Research Council (grant no 421-2012-892. We thank Demographic Data
Base for giving us access to data and Jonas Helgertz, Department of
Economic History, Lund University, for providing us with scripts for
creating HISCLASS from Swedish Census data.

>>>>

\bibliographystyle{apalike}
%\bibliographystyle{plainnat}

@@

\title{Long-term perspectives of social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Health and Mortality in Arctic Populations'. The Ninth International
    Congress on Arctic Social Sciences,
    Umeå,
    June 8--12, 2017}}

@@

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken in
this paper. The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013. The outcome is all
cause mortality and cause specific mortality for the categories
cardiovascular diseases and cancers. The analysis of inequality is based on
three social classes, reorganized from {\sc Hisclass}.

@@

The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf of Bothnia. This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century. The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s. One living long in memory was the famine of 1867–-1868 \citep{segb14}. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures. The inland was particularly remote and isolated. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden. Diseases reached to places that earlier were spared.

@@

In the data-set analysed here, all individuals 40 years and older ever being resident in either
of the regions are included. The data file contains the variables on social class ({\sc Hisclass}) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.

@@

In this paper we analyse mortality differences according to social class based on the classification scheme {\sc Hisclass} \citep{lema11}. The only type of socio-economic information available both in Linnaeus and Poplink is occupations. Thus, analysing according to social class is apart from being a scientific choice, also a practical one. Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid. Thus, the social class in the census 1990 is considered to be kept until 2001. This strategy is also used for other cases of missing social class. Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65. For the Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided to us by Jonas Helgertz, Lund University.

@@

{\sc Hisclass}, the classification system used here, is a "\ldots HISCO-based
historical international social class scheme" \citep{lema11, lema02}. The different classes in {\sc Hisclass} represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector. Implicitly it reflects large differences in what is in the forefront of other
categorisations, as for example access to economic and other resources, status, social power et cetera. We have merged the original 12 classes in {\sc Hisclass}, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:

\begin{enumerate}
\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals and higher white collar.
\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including farm workers.
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

@@

\subsection{Cause of death}
In the present study, we analyse mortality in the two most common disease
groups in the ages we are investigating---cardiovascular diseases and
cancers. National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year. Classification systems have however changed. Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system. In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.

@@

The important explanatory variable, {\tt hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\tt hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are presented \emph{graphically}. Tables with results from the different models are presented at the end of the article, but are not commented on in detail. The comments on the results are primarily based on the figures and focuses on the central variable in our study---{\tt hisclass}. As can be seen from the tables, most variables in the model are significant at 5 per cent level.

@@

\section{The effect of {\sc Hisclass}, 1851--1950, 1961--2013}

@@

From Table~\ref{tab:cumswtab} it is obvious that the statistical significance is strong.
The most interesting feature is the three-way interaction with {\tt sex}, {\tt period}, and {\tt hisclass}.

@@

As before, models are fitted separately for sex and each ten-year time period, from  1901 to
2013. We stratify on {\tt hisclass} and include the covariates {\tt urban} (TRUE or FALSE) and 
{\tt civst} (marital status). The cumulative
hazards at the end of the age period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of graphically illustrating the development over time.

@@

While we assume clear social differences in cardiovascular diseases, this is not necessarily an obvious assumption when it comes to cancers. Different forms of cancers have very different origin and different pathways to disease and death. The social pattern of higher mortality in classes with less power and accordingly less life chances are considered to be less clear when it comes to cancers. In our analyses, there is a risk that cancer mortality is under-estimated during the first half of the 20th century, illustrated by the already mentioned large proportion of unknown or vague causes at that time. We believe however that cancers were better diagnosed than cardiovascular deaths. This is supported by the stability in proportion of cancer deaths between the two first periods, 12.5 per cent 1901--1925 and 13.5 per cent for the period 1926--1950.

@@

Figure~\ref{fig:12cumsw} shows relative mortality levels according to {\sc Hisclass} during the different studied periods for the age group 40--64. Cancer mortality among the élite women was lower than that for the other classes during the complete studied period, while working class women had the highest levels during almost all periods. Consequently, we find a more consistent advantage of higher class in cancers compared to cardiovascular diseases for women. We do however not find the same pattern for men. It is difficult to identify any general features of 20th century male cancer mortality when it comes to social differences.

@@

See Table~\ref{tab:tableeducheart}.

@@

See Table~\ref{tab:tableeduccancer}.

@@

<<tableincprheart,results='asis'>>=
ltx(fit, dr = dr, caption = "Effect of income on cardiovascular mortality in ages 40-89, 1986-2005.", 
    label = "tab:tableincprheart", digits = 4)
@

See Table~\ref{tab:tableincprheart} for the effect of income on cardiovascular death.

@@

See Table~\ref{tab:tableincprcancer} for the income effect on cancer death.

@@

The authors of this paper have scrutinised different aspects of the social
pattern of mortality in historical northern Sweden for some time, focusing on
different regions, different age groups and different time periods \citep{edv92, edv04, edli11, edbr12}. When it comes to infant and child mortality pre---and 
early---transitional time, the expected socio-economic differences was often but not always confirmed \citep{edv92, edv04}, something that has been found in other studies as well \citep{bredro02, bube01, mol17}. Studies based on microdata on the adult population in history have been rare, mainly due to lack of adequate data. The results from such studies show other patterns than that of infant and child mortality. The expected social gradient in
survival, meaning that people from higher social classes always had an
advantage, was in many cases, particularly among adult men, not found
despite the well-documented socio-economic differences \citep{edli11, edbr12}. There were however large differences between men and women and for different age groups. These earlier studies have suggested that distinct and consistent differences to the advantage of higher classes developed somewhere during and probably quite early in the 20th century \citep{bevp11}. Until now we have however not been able to investigate the continuous development on mortality differences among the adult and elderly
population due to no available data. This is the first time it has been possible, and the results continue to surprise us. The consistent advantage for higher social classes do not appear until quite late in the regions we have analysed. The social pattern of female mortality are in line to what is
expected, but a social gradient in male adult mortality seems to be a recent phenomenon. Furthermore, this transition and the class pattern of mortality differ depending on age group. Inequality in mortality appear clearer for the population at working age, while the differences are smaller among the elderly.

@@

In this paper we have analysed social inequality in health from the
perspective of social classes using {\sc Hisclass}. However, there are
alternative ways to analyse social inequality, and we cannot exclude that
such analyses gives other results or at least would modify the image
presented here. In many modern studies categorisations based on income, wealth
or education are common for studies of inequality. They represent different dimensions of 
inequality. One problem is however that reliable data on income and
education are rarely available in historical population databases. 
As a test, we have included analyses of the associations of
education and income to mortality for recent decades. For women as
well as for men, for both age groups and for all-cause mortality as well as
cause-specific mortality we find consistent differences with higher
educational level and more income leading to lower mortality, thus basically confirming what we find in the class analysis. We do not know if education and income are equally important in the past.

@@

This work was undertaken within the Centre for Demographic and Ageing
Research (CEDAR) at Umeå University and with support from the Swedish
Research Council (VR) (grant No. 421-2012-892) and the project 
\emph{Old age, health, and inequality: Mortality from a life course perspective in Sweden 1986--2009}, 
 financed by the Swedish Research Council (grant No. 421-2012-892). We thank The Demographic Data
Base for giving us access to data and Jonas Helgertz, Department of
Economic History, Lund University, for providing us with scripts for
creating {\sc Hisclass} from Swedish Census data.
